---
title: "Lost Potential: Exploring Youth Unemployment in the Global Labour Force"
subtitle: "From Cradle to Career: The Quirks and Queries of Youth Unemployment"
author: "Risikat Sulaimon"
date: "2024-04-10"
format:
  html:
    theme: sandstone
    toc: true
    toc-title: "The Unemployment Safari!"
    toc_float: true 
execute:
  echo: false
  warning: false
  message: false
---
```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(plotly)
library(ggplot2)
library(countrycode)
library(scales)
library(dplyr)
library(viridis)
library(maps)

# data
unicef_indicator_1_5_ <- read_csv("C:/Users/rissu/Downloads/unicef_indicator_1 (5).csv")
unicef_metadata_5_ <- read_csv("C:/Users/rissu/Downloads/unicef_metadata (5).csv")
#Transformations
Data_join <- full_join(unicef_indicator_1_5_, unicef_metadata_5_, by = join_by(country==country, time_period==year))
```

## Introduction {style="color:lightpink;"}

*Using information from UNICEF indicators, this report explores the intricate world of youth unemployment, concentrating on the **Percentage of adolescents (aged 15-19 years) in the labor force who are unemployed**.* This study aims to explore the complex issues young people encounter while navigating the global labor market by utilizing UNICEF megadata. Policymakers, educators, and stakeholders must all grasp the underlying causes and consequences of the rising youth unemployment rates in many areas. This report looks at the interactions between gender dynamics, countries worldwide, continents, life expectancy and population in an effort to shed light on the obstacles that youth employment faces and investigate viable solutions.


##  A Journey Through Data Charts
With the utilization of UNICEF data, I have crafted the following charts below to glean insights into the status of unemployed adolescents aged 15 to 19 globally:

- **World maps**
- **Time series**
- **Scatter plot**
- **Bar chart**
- **Pie chart**
- **Histogram**
- **Stacked bar chart**

```{r}
#| fig-cap: "World map based on UNICEF indicator data, depicting the geographical distribution of unemployed adolescents juxtaposed with their life expectancy offering valuable insights into regional variations and highlighting areas for targeted interventions and policy initiatives"
#| code-fold: true
#| label: fig-map1
library(ggplot2)
library(dplyr)

# Load necessary packages
if (!requireNamespace("maps", quietly = TRUE)) {
  install.packages("maps")
}
library(maps)

# Check if 'map_data' function is available
if (!exists("map_data")) {
  stop("The 'map_data' function from the 'maps' package is not available.")
}

# Attempt to create 'map_world' object
tryCatch({
  map_world <- map_data("world")
}, error = function(e) {
  stop("Error creating 'map_world' object: ", conditionMessage(e))
})

# Check if 'Data_join' dataset is available
if (!exists("Data_join")) {
  stop("The 'Data_join' dataset is not available.")
}

# Check if required columns are present in 'Data_join'
if (!all(c("country", "Life expectancy at birth, total (years)") %in% names(Data_join))) {
  stop("Required columns are missing in 'Data_join'.")
}

# Attempt to join datasets
tryCatch({
  map_data_join <- full_join(Data_join, map_world, by = c("country" = "region"))
}, error = function(e) {
  stop("Error joining datasets: ", conditionMessage(e))
})

# Check if join was successful
if (!exists("map_data_join")) {
  stop("Joining datasets failed.")
}

# Plot the map
ggplot(map_data_join, aes(x = long, y = lat, group = group, fill = `Life expectancy at birth, total (years)`)) +
  geom_polygon(color = "black") +  
  scale_fill_viridis(name = "Life Expectancy", option = "plasma", na.value = "gray50", guide = "legend") + 
  labs(title = "World Map: Life Expectancy", x = NULL, y = NULL) + 
  theme_minimal()
```
```{r}
#| fig-cap: "World map based on UNICEF indicator data, illustrating the geographical distribution of unemployed adolescents alongside their life expectancy in 1960, offering a snapshot of historical conditions and trends. This visualization serves as a comparison to the recent life expectancy world map @fig-map1 to discern changes over time and assess the evolution of regional variations, providing insights for targeted interventions and policy initiatives."


#World Map

#| #| label: fig-map2
data_join_1960 <- Data_join %>%
  filter(time_period == "1960")
map_data_join_1960 <- full_join(data_join_1960, map_world, by = c("country" = "region"))
ggplot(map_data_join_1960, aes(x = long, y = lat, group = group, fill = `Life expectancy at birth, total (years)`)) +
  geom_polygon(color = "black") +  # Add black borders to polygons
  scale_fill_gradient(name = "Life Expectancy", low = "lightblue", high = "darkblue") +  # Use gradient fill colors
  labs(title = "World Map: Life Expectancy in 1960", x = NULL, y = NULL) +  # Add title and remove axis labels
  theme_minimal()
```
```{r}
#| fig-cap: "Time series showcasing the observed values of unemployed adolescents worldwide. This visualization provides insights into the distribution and magnitude of youth unemployment across various countries and regions. By examining the observed values over time, we gain valuable insights into the trends and patterns of youth unemployment on a global scale, informing policymakers and stakeholders about the challenges faced by young people in accessing employment opportunities and contributing to the economy"
#| label: fig-timeseries1


 #Time Series
Data_join$continent <- countrycode(Data_join$country, "country.name", "continent")
timeseries_plot_2 <- Data_join %>%
  ggplot() +
  aes(x = time_period, y = `Life expectancy at birth, total (years)`, color = continent) +
  geom_line() +
  labs(title = "**Time Series of Life Expectancy by Continent**",
       x = "Time Period",
       y = "Life Expectancy at Birth (years)") +
  scale_color_brewer(palette = "Set1") +  # Using Set1 color palette for continents
  theme_minimal()  # Use a minimal theme for better readability



ggplotly(timeseries_plot_2)
```
```{r}
#| fig-cap: "Scatter plot illustrating the relationship between observed values and life expectancy, with bubble size representing the population size of unemployed adolescents worldwide. The red regression line highlights the overall trend in the data. The data is sourced from UNICEF and provides insights into the global distribution of unemployment among adolescents and its potential impact on life expectancy"
#| label: fig-scatterplot1
# Create the scatter plot
Scatter_plot_1 <- ggplot(Data_join, aes(x = obs_value, y = `Life expectancy at birth, total (years)`, size = `Population, total`)) +
  geom_point(alpha = 0.6, color = "darkblue") +  # Scatter plot points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Red regression line
  labs(
    x = "Observed Value",
    y = "Life Expectancy (Years)",
    title = "Life Expectancy vs. Observed Value",
    size = "Population"
  ) + 
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels with commas
  scale_size_continuous(range = c(2, 10)) +  # Adjust size range for better visibility
  theme_minimal() +  # Use minimal theme for simplicity
  theme(
    text = element_text(family = "sans"),
    plot.title = element_text(hjust = 0.5),  # Center plot title
    axis.title = element_text(size = 10),  # Adjust axis title size
    legend.position = "none"  # Remove legend
  )

# Convert ggplot to plotly object
ggplotly(Scatter_plot_1)

```
```{r}
#| fig-cap: "Bar chart illustrating the evolution of average life expectancy per continent from 1960 to 2021, alongside the prevalence of unemployed adolescents. This visualization offers a comprehensive view of how life expectancy has changed over time across different continents, while also highlighting the challenges of youth unemployment. By juxtaposing these two indicators, we can assess the impact of socioeconomic factors on health outcomes and identify regions where targeted interventions are needed to address both unemployment among young people and improve overall population health"
#| label: fig-barchart1
 # BAR CHART 1960-2021
  library(ggplot2)
  library(dplyr)
  library(countrycode)
  
  # Create a function to assign continents based on countries
  assign_continent <- function(country) {
    # Use countrycode package to get continents from countries
    continent <- countrycode(country, "country.name", "continent")
    return(continent)
  }
  
  # Apply the function to create a new column 'continent'
  Data_join <- Data_join %>%
    mutate(continent = assign_continent(country))
  
  # Filter data for years 1980, 2000, 2010, and 2020
  Data_filtered <- Data_join %>% 
    filter(time_period %in% c(1980, 2000, 2010, 2020))
  
  
  Data_filtered %>%
    group_by(continent, time_period) %>%
    summarise(avg_life_exp = mean(`Life expectancy at birth, total (years)`, na.rm = TRUE), .groups = "drop") %>%
    
   
    ggplot(aes(x = reorder(continent, avg_life_exp), y = avg_life_exp, fill = continent)) +
    geom_bar(stat = "identity") +
    
   
    facet_wrap(~time_period) +
    
   
    labs(
      x = "",
      y = "Average Life Expectancy",
      fill = "Continent",
      title = "Evolution of Average Life Expectancy per Continent for 1960-2021"
    ) +
    
  
    theme_classic() +
    theme(
      text = element_text(family = "sans"),
      axis.text.x = element_blank()
    )
```


```{r}
#| fig-cap: "In a pie chart depicting global youth unemployment, Europe emerges as the surprising epicenter, dominating the landscape. This revelation, drawn from UNICEF data, sparks inquiry into the underlying factors and prompts action to address the concentrated challenges facing unemployed adolescents in the region.."
#| label: fig-piechart1
#pie chart
  
  library(ggplot2)
  library(scales)

  continent_counts <- Data_join %>%
    group_by(continent) %>%
    summarise(total_obs = sum(obs_value, na.rm = TRUE))
  
  continent_counts$percent <- continent_counts$total_obs / sum(continent_counts$total_obs) * 100
  

  continent_counts$percent_rounded <- round(continent_counts$percent, 2)
  

  pie_chart <- ggplot(continent_counts, aes(x = "", y = percent, fill = continent)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    geom_text(aes(label = paste0(percent_rounded, "%")), position = position_stack(vjust = 0.5), color = "white", size = 4) +
    labs(title = "Distribution of Unemployed Adolescents by Continent", fill = "Continent") +
    theme_minimal() +
    theme(legend.position = "right") +
    scale_fill_brewer(palette = "Set3") +
    guides(fill = guide_legend(title = "Continent")) +
    scale_y_continuous(labels = percent_format()) +
    theme(axis.text.x = element_blank()) +  # Hide x-axis labels
    theme(axis.text.y = element_blank()) +  # Hide y-axis labels
    theme(plot.title = element_text(hjust = 0.5))  # Center plot title
  
  pie_chart
  
```
```{r}
#| fig-cap: "Exploring the Distribution: Histogram depicting the frequency distribution of observed values among unemployed adolescents. This visualization sheds light on the spread and concentration of unemployment rates among adolescents, providing valuable insights for targeted interventions and policy formulation"
#| label: fig-histogram1

 #Histogram
  
  library(ggplot2)
  

  ggplot(Data_join, aes(x = obs_value)) +
    geom_histogram(binwidth = 5, fill = "#66c2a5", color = "#1f78b4", alpha = 0.8) +
    labs(
      x = "Unemployed Adolescent",
      y = "Frequency",
      title = "Distribution of Unemployed Adolescents",
      subtitle = "Histogram showing the frequency of unemployed adolescents"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12, face = "bold")
    )
 

```
```{r}
#| fig-cap: "This stacked bar chart illustrates the distribution of unemployed adolescents by sex across different continents. Notably, it reveals a higher prevalence of female unemployment compared to males, shedding light on significant gender disparities in the global labor force. Understanding these imbalances is crucial for devising targeted interventions aimed at fostering gender equity and promoting equal opportunities worldwide"
#| label: fig-stackedbarchart1
 # Stacked bar chart by continent
  
  install.packages("countrycode")
  library(countrycode)

unicef_indicator_1_5_$continent <- countrycode(unicef_indicator_1_5_$country, "country.name", "continent")
continent_data <- unicef_indicator_1_5_ %>%
  group_by(continent, sex) %>%
  summarise(total_obs = sum(obs_value))


ggplot(continent_data, aes(x = continent, y = total_obs, fill = sex)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Continent",
    y = "Number of Unemployed Adolescents",
    fill = "Sex",
    title = "Distribution of Unemployed Adolescents by Sex Across Continents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



 #stacked bar chart

library(ggplot2)

ggplot(unicef_indicator_1_5_, aes(x = current_age, fill = sex)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("Male" = "skyblue", "Female" = "pink")) +
  labs(
    x = "Current Age",
    y = "Number of Unemployed Adolescents",
    fill = "Sex",
    title = "Stacked Bar Chart of Unemployed Adolescents by Sex and Current Age"
  ) +
  theme_minimal()

```
## Conclusion and Recommendation {style="color:lightpink;"}

## <u>Recommendations</u> {style="color: gold;"}
- **Invest in Education:** Allocate resources towards education and skill development programs to equip young people with the necessary tools to succeed in the modern job market.
- **Promote Innovation:** Foster an environment that encourages innovation and entrepreneurship, creating opportunities for young people to thrive and contribute to economic growth.
- **Create Job Opportunities:** Implement policies and initiatives aimed at job creation, particularly targeting industries with high potential for youth employment.
- **Address Gender Disparities:** Develop strategies to address gender imbalances in the workforce, ensuring equal opportunities for all young people.
- **Global Collaboration:** Foster international collaboration and partnerships to share best practices and resources for combating youth unemployment on a global scale.


## <u>Conclusion</u> {style="color: green;"}
The report dismantles the intricate tapestry of socio-economic factors interwoven within the global youth unemployment crisis. While familiar themes like gender disparities and educational inequalities echo through its pages, disparities persist across continents. To effectively tackle this challenge, policymakers must prioritize investments in education and skill development, nurturing an environment conducive to innovation and job creation. By harmonizing efforts and embracing collaborative solutions, we can compose a future where the talents and aspirations of our youth resonate harmoniously with the demands of the modern labor market.

::: {.callout-tip title="More information" style="color: black;"}
*See [https://data.unicef.org/](https://data.unicef.org/) and [https://data.unicef.org/indicator-profile/](https://data.unicef.org/indicator-profile/) for more information*
:::

